---
title: "Analysing countries of the world using ggplot2"
author: "David Ju Nyambe"
date: "1/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

## Questions to be answered

1. Top 30 countries with the highest populations
2. Top 30 countries with the largest area
3. Birthrate and deathrate of 50 countries with the highest literacy levels
4. Birthrate and deathrate of 50 countires with the lowest literacy
5. GDP by region
6. GDP by birthrate
7. Highest gdp per Capita
8. Average regional GDP


The countries of the world dataset is used for this publication. It is available for download here [Kaggle](https://www.kaggle.com/fernandol/countries-of-the-world).

## Loading the libraries
```{r loading_libraries,results='hide',message=FALSE}
install.packages("ggplot2",repos = "http://cran.us.r-project.org")
install.packages("tidyverse",repos = "http://cran.us.r-project.org")
install.packages("janitor",repos = "http://cran.us.r-project.org")
library("tidyr")
library("stringr")
library("dplyr")
library("janitor")
library(ggplot2)
```

## Loading the countries of the world data set

```{r countries}
countries<-read.csv("countries of the world.csv",sep =',',header = TRUE)
```

## Data Wrangling

The str function is used to take a glimpse at the countries dataset using glimpse() from the dplyr library
```{r processing}
glimpse(countries)
```
From a simple glimpse of the dataset several observations are made that will need to be corrected these things are   

* Uppercase letters and spaces in some of the variables(Column names).
* Some observations have "," instead of "." .
* The region  variable rows have white spaces.
* Some variables have wrong data types.

using the clean_names() from the janitor library, all uppercase letters will be converted to lowercase and all spaces will be converted to underscores(_) in the variable names.
```{r replacing_whitespaces}
countries<-countries %>% clean_names()
names(countries)
  
```


A function add_point is created that will be used to replace "," with "." in the observations.
```{r addpoint}
  add_point <-  function(x){
  str_replace(x, ',','.')
  }

countries<- countries %>% mutate_all(add_point)
glimpse(countries)

```

using stringr's str_trim(), the white spaces in region variables rows are removed
```{r removing_whitespaces}
  countries <- countries %>% mutate(region=str_trim(region,side=c("both")))

```

Using the mutate function from the dplyr library, the variables(columns) are converted to their appropriate datatype

```{r datatypes}
    countries<-countries %>% mutate(country=as.factor(country),
                                region=as.factor(region),
                                population=as.numeric(population),
                                climate=as.factor(climate),
                                area_sq_mi=as.numeric(area_sq_mi),
                                pop_density_per_sq_mi=as.numeric(pop_density_per_sq_mi),
                                coastline_coast_area_ratio=as.numeric(coastline_coast_area_ratio),
                                net_migration=as.numeric(net_migration),
                                infant_mortality_per_1000_births=as.numeric(infant_mortality_per_1000_births),
                                gdp_per_capita=as.numeric(gdp_per_capita),
                                literacy=as.numeric(literacy),
                                phones_per_1000=as.numeric(phones_per_1000),
                                arable=as.numeric(arable),
                                crops=as.numeric(crops),
                                other=as.numeric(other),
                                birthrate=as.numeric(birthrate),
                                deathrate=as.numeric(deathrate),
                                agriculture=as.numeric(agriculture),
                                industry=as.numeric(industry),
                                service=as.numeric(service))
```

A check is performed to see if there are any missing values
```{r missing}
glimpse(countries)
colSums(is.na(countries))
```

It is observed that the dataset contains some missing values , therefore all missing values are replaced with the mean of the variable (column).
```{r filter_out_missing_values,results='hide'}
means_to_replace<- select_if(ungroup(countries), is.numeric) %>%
 colMeans(na.rm=T)
format_mean<-function(.x){
  ifelse(is.na(.x),as.numeric(format(round(mean(.x, na.rm = TRUE),3),nsmall=3)), as.numeric(.x))
  
}
 countries<-bind_cols(select(ungroup(countries),!is.numeric),
          select_if(ungroup(countries), is.numeric) %>% 
            mutate_all(format_mean)
          ) 
```

Check if missing values have been replaced with the mean
```{r replaced_mean}
colSums(is.na(countries))
```
## Answering the questions using ggplot2

### 1. Top 30 countries with the highest population 
```{r highest_population}
countries_with_highest_populations<- countries %>% select(country,population) %>%mutate(population=as.numeric(format(round((population/10^9),2),nsmall=2))) %>% arrange(desc(population)) %>% head(n=10)

ggplot(data=countries_with_highest_populations,mapping=aes(x=reorder(country,-population),y=population)) +
  geom_bar(fill="steel blue",stat="identity")+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_text(aes(label=population), position=position_dodge(width=0.9), vjust=-0.25,size=2.5) +labs(title="Top 30 countries with the highest population",x ="Country",y="Population (in Billions)")



```

### 2. Top 30 Countries with the largest area
```{r largest_area}
countries_with_largest_area<- countries %>% select(country,area_sq_mi) %>%mutate(area_sq_mi=as.numeric(format(round((area_sq_mi/10^4),2),nsmall=2))) %>% arrange(desc(area_sq_mi)) %>% head(n=10)

ggplot(data=countries_with_largest_area,mapping=aes(x=reorder(country,-area_sq_mi),y=area_sq_mi)) +
  geom_bar(fill="steel blue",stat="identity")+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_text(aes(label=area_sq_mi), position=position_dodge(width=0.9), vjust=-0.25,size=2.5) +labs(title="Top 10 countries with the largest Area",x ="Country",y="Area (per 100,0000m)") 
```


### 3. Leading sectors for top 10 countries with the highest G.D.P in US Dollars
```{r leading_sectors}
leading_sectors_for_countries_with_highest_gdp<- countries %>% select(country,gdp_per_capita,agriculture,industry,service) %>%mutate(leading_sector=case_when(agriculture > industry & agriculture > service ~"agriculture",
  industry > agriculture & industry > service ~ "industry",
  service > agriculture & service > industry ~"service")) %>% 
  arrange(desc(gdp_per_capita)) %>% head(n=10)

ggplot(data=leading_sectors_for_countries_with_highest_gdp,mapping=aes(x=reorder(country,-gdp_per_capita),y=gdp_per_capita,fill=leading_sector)) +
  geom_bar(stat="identity")+
   scale_y_continuous(labels = scales::dollar) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_text(aes(label=gdp_per_capita), position=position_dodge(width=0.9), vjust=-0.25,size=2.5) +labs(title="Leading sectors for top 10 countries with the higest GDP in US Dollars",x ="Country",y="G.D.P per capita in US Dollars ($)") 
```


## 4. Leading sectors in regions based on their G.D.P in U.S Dollars($)
```{r leading_sectors_in_regions}
leading_sectors_for_regions_based_on_gdp<- countries %>% select(region,gdp_per_capita,agriculture,industry,service)%>% group_by(region) %>%
  mutate(leading_sector=case_when(agriculture > industry & agriculture > service ~"agriculture",
  industry > agriculture & industry > service ~ "industry",
  service > agriculture & service > industry ~"service"),gdp=sum(gdp_per_capita)) %>% 
  arrange(desc(gdp)) %>% select(region,gdp,leading_sector) %>% distinct(region,.keep_all=TRUE)

ggplot(data=leading_sectors_for_regions_based_on_gdp,mapping=aes(x=reorder(region,-gdp),y=gdp,fill=leading_sector)) +
  geom_bar(stat="identity")+
   scale_y_continuous(labels = scales::dollar) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_text(aes(label=gdp), position=position_dodge(width=0.9), vjust=-0.25,size=2.5) +labs(title="Leading sectors for top 10 countries with the higest GDP in US Dollars",x ="Region",y="G.D.P per capita in US Dollars ($)") 
```


## 5. Top 10 countries with the highest infant mortality rate

```{r higest_infant_mortality}
countries_with_highest_infant_mortalities<- countries %>% select(country,infant_mortality_per_1000_births) %>% arrange(desc(infant_mortality_per_1000_births)) %>% head(n=10)

ggplot(data=countries_with_highest_infant_mortalities,mapping=aes(x=reorder(country,-infant_mortality_per_1000_births),y=infant_mortality_per_1000_births,)) +
  geom_bar(stat="identity",fill="steel blue")+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_text(aes(label=infant_mortality_per_1000_births), position=position_dodge(width=0.9), vjust=-0.25,size=2.5) +labs(title="Top 10 countries with the highest infant mortality rates",x ="Country",y="Mortality per 1000 births")

```

## 6. Countries with the lowest literacy levels
```{r higest_literacy_levels}
countries_with_highest_literacy_levels<- countries %>% select(country,literacy) %>% arrange(literacy) %>% head(n=10)

ggplot(data=countries_with_highest_literacy_levels,mapping=aes(x=reorder(country,-literacy),y=literacy)) +
  geom_bar(stat="identity",fill="steel blue")+  
  scale_y_continuous(labels =scales::percent_format(scale = 1)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_text(aes(label=literacy), position=position_dodge(width=0.9), vjust=-0.25,size=2.5) +labs(title="Top 10 countries with the lowest literacy levels",x ="Country",y="literacy level")

```

## 7. Birthrate and Deathrate of  top 10 countries with high G.D,P

```{r birthrate_by_gdp}
countries_rates_with_high_gdp <- countries %>% select(country,gdp_per_capita,birthrate,deathrate) %>%
  pivot_longer(c(deathrate,birthrate),names_to="rate") %>%
  arrange(desc(gdp_per_capita)) %>% 
head(n=20)

ggplot(data=countries_rates_with_high_gdp,mapping=aes(x=reorder(country,-value),value,fill=rate)) +
  geom_bar(stat="identity",position="dodge")+
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
 labs(title="Birthrate  and deathrate of countries with high G.D.P per capita",x ="Country",y="birthtate and deathrate")

```

## 8. Birthrate and Deathrate of top 10 countries  with low G.D,P

```{r hrate_by_low_gdp}
countries_rates_with_low_gdp <- countries %>% select(country,gdp_per_capita,birthrate,deathrate) %>%
  pivot_longer(c(deathrate,birthrate),names_to="rate") %>%
  arrange(gdp_per_capita) %>% 
head(n=20)

ggplot(data=countries_rates_with_low_gdp,mapping=aes(x=reorder(country,-value),value,fill=rate)) +
  geom_bar(stat="identity",position="dodge")+
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
 labs(title="Birthrate  and deathrate of countries with high G.D.P per capita",x ="Country",y="brithtrate and deathrate")

```
